pyTopoComplexity: A Python package for topographic complexity analysis

### 2024-08-23 JOSS/paper.rmd
Add files via upload

---
title: 'pyTopoComplexity: A Python package for topographic complexity analysis'
tags:
  - Python
  - topographic complexity
  - surface roughness
  - two-dimensional continuous wavelet transform
  - fractal dimension
  - rugosity index
authors:
  - name: Larry Syu-Heng Lai
    corresponding: true
    orcid: 0000-0003-4589-9729 
    affiliation: "1"
  - name: Adam M. Booth
    orcid: 0000-0002-7339-0594   
    affiliation: "2"
  - name: Alison R. Duvall
    orcid: 0000-0002-7760-7236 
    affiliation: "1"
  - name: Erich Herzig
    orcid: 0000-0002-3898-5971 
    affiliation: "1"
affiliations:
 - name: Department of Earth and Space Sciences, University of Washington, Seattle, Washington, USA
   index: 1
 - name: Department of Geology, Portland State University, Portland, Oregon, USA
   index: 2
date: 18 August 2024
bibliography: paper.bib
---

@@

**pyTopoComplexity** is a Python package that provides a computationally efficient and customizable 
implementation of three methods for quantifying topographic complexity. These methods include 
two-dimensional continuous wavelet transform (2D-CWT) analysis, fractal dimension estimation, and 
rugosity index calculation across various spatial scales. This package addresses the scarcity of 
open-source software for these sophisticated methods, which are crucial in modern terrain analysis, 
and facilitates data comparison and reproducibility. In the [software respository](https://github.com/GeoLarryLai/pyTopoComlexity.git), 
we also include a Jupyter Notebook file that integrates components from the Python-based 
surface-process modeling platform **Landlab** [@Hobley2017]. This allows researchers to 
simulate the smoothing of topography over time through terrestrial nonlinear hillslope 
diffusion processes. By combining these features, **pyTopoComplexity** advances the toolset 
available to researchers for measuring and simulating the time-dependent persistence of 
topographic complexity signatures against environmental forces on terrain surfaces.

@@

In recent years, several advanced methods for quantifying topographic complexity have been developed, 
including two-dimensional continuous wavelet transform (2D-CWT) analysis [@Booth2009; @Berti2013], 
fractal dimension estimation [@Taud2005; @Robbins2018; @PardoIguzquiza2020], and rugosity index 
calculation [@Jenness2004; @DuPreez2015]. These methods are considered more effective for terrain analysis 
tasks compared to conventional morphological metrics such as variations in local slope and relief. 
Despite their importance, comprehensive publicly available tools that incorporate these advanced 
methods for studying topographic complexity are lacking. Common open-source geospatial analysis software, 
such as QGIS [@QGIS_software], GRASS GIS [@GRASS_GIS_software], and WhiteboxTools [@Lindsay2016], only 
implement basic conventional methods, limiting the reproducibility and comparability of these newer 
approaches. Although some specialized programs for calculating the rugosity index exist [@Walbridge2018; @Benham2022], 
they have been limited to marine bathymetric studies and involve various mathematical choices and designs.

To address this gap, we have developed an open-source Python toolkit called **pyTopoComplexity**, which provides 
computationally efficient and easily customizable implementations of three modules for performing and 
visualizing the results of 2D-CWT, fractal dimension, and rugosity calculations (Table 1\autoref{tab:1}). 
This toolkit can detect the grid spacing and unit of the projected coordinate system (acceptable in meters, 
U.S. survey feet, and international feet) from the input raster DTM file (GeoTIFF format) and 
automatically conduct unit conversions in necessary calculation steps to ensure data consistency and 
reproducibility. Results at nodes affected by edge effects due to no-data values outside the input raster 
will be removed by default. Users can define the suitable spatial scale to match their research purposes 
and choose computational approaches (e.g., chunk processing, faster mathematical approximations) to 
optimize performance (see details in the **Methods and features overview** section).

| Modules  | Classes | Method Descriptions | References |
| ------------- | ------------- | ------------- | ------------- |
| pycwtmexhat.py | CWTMexHat | Quantifies the wavelet-based curvature of the land surface using two-dimensional continuous wavelet transform (2D-CWT) with a Mexican Hat wavelet | @Booth2009; @Booth2017 |
| pyfracd.py | FracD | Conducts fractal dimension analysis on the land surface using variogram procedures | @Wen1997; @PardoIguzquiza2020|
| pyrugostiy.py | RugosityIndex | Calculates the rugosity index of the land surface | @Jenness2004; @DuPreez2015 |
Table: Table 1 \label{tab:1}: Modules contained in the **pyTopoComplexity** package.

Each module of the **pyTopoComplexity** is provided with a corresponding [example Jupyter Notebook file](https://github.com/GeoLarryLai/pyTopoComlexity/tree/main/example) 
for usage instructions, using the Light Detection and Ranging (LiDAR) DTM data of a deep-seated landslide that 
occurred in 2014 in the Oso area of the North Fork Stillaguamish River valley, Washington State, USA 
[@WashingtonGeologicalSurvey2023]. In the software repository, we also include an additional Jupyter Notebook file 
[**nonlineardiff_Landlab.ipynb**](https://github.com/GeoLarryLai/pyTopoComlexity/blob/main/example/nonlineardiff_Landlab.ipynb), 
which allows researchers to simulate the smoothing of topography over time via terrestrial nonlinear hillslope 
diffusion processes [@Roering1999]. This is achieved by employing the `TaylorNonLinearDiffuser` module from the 
`terrainBento` Python package [@Barnhart2019] and running the simulation in the **Landlab** environment [@Hobley2017].

@@

The Mexican hat wavelet is proportional to the second derivative of a Gaussian envelope, with its Fourier wavelength ($\lambda$) 
ependent on the grid spacing ($\Delta$) of the input DTM raster. The wavelet function $\psi$ is scaled according to the 
wavelet scale parameter $s$ and the grid spacing $\Delta$, ensuring that the resultant wavelet coefficient $C$ signifies 
concave and convex landforms corresponding to the wavelet scale $s$. Users can define the value of $\lambda$ in meters as the 
targeted spatial scale for landform roughness analysis, and the **pycwtmexhat.py** module will automatically compute the wavelet 
scale $s$ based on the grid spacing ($\Delta$) of the input raster file (Figure 1)\autoref{fig:1}.

@@

, where $TD$ is the topological dimension in the Euclidean space of the fractional Brownian motion. For one-dimensional fractional Brownian 
motion, $TD = 1$; thus, the fractal dimension of the two-dimensional surface $(FD)_2^*$ can be estimated as the average fractal dimension of 
the one-dimensional profiles $(FD)_1^*$:

@@

## Rugosity index calculation

The **pyrugosity.py** module in **pyTopoComplexity**  measures the rugosity index of the land surface, which is widely used to assess structural 
complexity of the topography. Such method has been applied in classifying seafloor types by marine geologists and geomorphologist, small-scale 
hydrodynamics by oceanographers, and studying available habitats in the landscape by ecologists and coral biologists [@Lundblad2006; @Wilson2007].

The rugosity index is determined as the ratio of the contoured area (i.e., true geometric surface area) to the planimetric area within the square 
moving window, highlighting smaller-scale variations in surface height:

@@

This module adapts the Triangulated Irregular Networks (TIN) method from @Jenness2004 to approximate the contoured area as the sum of 
eight truncated-triangle areas. These triangles connect the central grids, four corner grids, and four grids at the middle points of the 
surrounding edges within the moving window. If no local slope correction is applied, the planimetric area is considered to be the horizontal 
planar area of the moving window, as described in @Jenness2004. Another approach considers slope correction where to the planimetric area is 
projected onto an plane of the local gradient [@DuPreez2015].

By definition, the rugosity index is as a minimum value of one (completely flate surface). Typical valuesrange of the conventional rugosity 
index (without slope correction) from one to three although larger values are possible in very steep terrains. The slope-corrected rugosity 
index, also called arc-chord ratio (ACR) rugosity index, could provide a better representation of local surface complexity (Figure 3)\autoref{fig:3}.

![Figure 3. Hillshade map of the 2014 Oso Landslide region (left), along with the calculated results of the Arc-Chord Ratio Rugosity Index (middle) 
and the conventional Rugosity Index (right), using the **pyrugosity.py** module. The calculations were performed with a designated 17 by 17 grid 
moving window size (grid spacing = 3 U.S. survey feet ≈ 0.9144 meters). \label{fig:3}](fig3-pyrugosity.png){ width=100% }

@@

The `nonlineardiff_Landlab.ipynb` notebook in the **pyTopoComplexity** package offers a sophisticated tool for simulating landscape evolution 
through nonlinear diffusion processes due to near-surface soil disturbances and downslope sediment creep [@Roering1999]. This tool runs the 
simulation in the **Landlab** environment (version >= 2.7) [@Hobley2017] with the `TaylorNonLinearDiffuser` module from the `terrainBento` 
Python package [@Barnhart2019]. The main simulation iteratively applies the nonlinear diffusion model to predict changes in surface 
elevation $z$ over time $t$:

@@

This notebook provides a comprehensive workflow that guides users through setting up, importing raster files, and running simulations. Since 
**Landlab** primarily handles DTM data in ESRI ASCII format, this notebook includes utility functions for converting raster files between GeoTIFF 
and ESRI ASCII formats. Users are required to specify the values for $S_c$, $K$, the length of each time step in years, and the final time to 
stop the simulation. The example included in the notebook uses LiDAR DTM data from the 2014 Oso Landslide [@WashingtonGeologicalSurvey2023], 
with parameter values provided in @Booth2017, in an attempt to reproduce the simulation results presented in that study (Figure 4\autoref{fig:4}).

![Figure 4. Hillshade map of the 2014 Oso Lanslide region and surface smoothing evolution over 15,000 years predicted by a 
nonlinear hillslope diffusion model used in the `pyrugosity` module, in attempt to reproduce the simulation results in 
@Booth2017.\label{fig:3}](fig4-smoothing.png){ width=100% }

@@

The development of **pyTopoComplexity** is part of a collaborative effort within the [Landslide subteam](https://cascadiacopeshub.org/team-1-landslides/) of the 
[Cascadia Coastlines and Peoples Hazards Research Hub (Cascadia CoPes Hub)](https://cascadiacopeshub.org), funded by the National Science Foundation Award ([#2103713](https://www.nsf.gov/awardsearch/showAward?AWD_ID=2103713&HistoricalAwards=false)). 
We extend special thanks to Dr. Eulogio Pardo-Igúzquiza, who generously shared his Fortran code for fractal dimension analysis used in his work 
@PardoIguzquiza2022b, which significantly inspired the development of the **pyfracd.py** module.

>>>>

---
title: 'pyTopoComplexity: A Python package for topographic complexity analysis'
tags:
  - Python
  - topographic complexity
  - surface roughness
  - two-dimensional continuous wavelet transform
  - fractal dimension
  - Rugosity Index
  - Terrain Position Index
authors:
  - name: Larry Syu-Heng Lai
    corresponding: true
    orcid: 0000-0003-4589-9729 
    affiliation: "1"
  - name: Adam M. Booth
    orcid: 0000-0002-7339-0594   
    affiliation: "2"
  - name: Alison R. Duvall
    orcid: 0000-0002-7760-7236 
    affiliation: "1"
  - name: Erich Herzig
    orcid: 0000-0002-3898-5971 
    affiliation: "1"
affiliations:
 - name: Department of Earth and Space Sciences, University of Washington, Seattle, Washington, USA
   index: 1
 - name: Department of Geology, Portland State University, Portland, Oregon, USA
   index: 2
date: 18 August 2024
bibliography: paper.bib
---

@@

**pyTopoComplexity** is a Python package that provides a computationally efficient and customizable 
implementation of four modern methods for quantifying topographic complexity. These methods include 
two-dimensional continuous wavelet transform (2D-CWT) analysis, fractal dimension estimation, Rugosity 
Index and Terrain Position Index calculations across various spatial scales. This package addresses 
the scarcity of open-source software for these sophisticated methods, which are crucial in modern 
terrain analysis, and facilitates data comparison and reproducibility. In the [software respository](https://github.com/GeoLarryLai/pyTopoComplexity.git), 
we also include a Jupyter Notebook file that integrates components from the Python-based 
surface-process modeling platform **Landlab** [@Hobley2017]. This allows researchers to 
simulate the smoothing of topography over time through terrestrial nonlinear hillslope 
diffusion processes. By combining these features, **pyTopoComplexity** advances the toolset 
available to researchers for measuring and simulating the time-dependent persistence of 
topographic complexity signatures against environmental forces on terrain surfaces.

@@

In recent years, several advanced methods for quantifying topographic complexity have been developed, 
including two-dimensional continuous wavelet transform (2D-CWT) analysis [@Booth2009; @Berti2013], 
fractal dimension estimation [@Taud2005; @Robbins2018; @PardoIguzquiza2020], and Rugosity Index 
calculation [@Jenness2004; @DuPreez2015]. These methods are considered more effective for terrain analysis 
tasks compared to conventional morphological metrics such as variations in local slope and relief. 
Despite their importance, comprehensive publicly available tools that incorporate these advanced 
methods for studying topographic complexity are lacking. Common open-source geospatial analysis software, 
such as QGIS [@QGIS_software], GRASS GIS [@GRASS_GIS_software], and WhiteboxTools [@Lindsay2016], only 
implement basic conventional methods, limiting the reproducibility and comparability of these newer 
approaches. Although some specialized programs for calculating the Rugosity Index exist [@Walbridge2018; @Benham2022], 
they have been limited to marine bathymetric studies and involve various mathematical choices and designs.

To address this gap, we have developed an open-source Python toolkit called **pyTopoComplexity**. This toolkit 
offers computationally efficient and easily customizable implementations for performing and visualizing the results 
of 2D-CWT, fractal dimension, and rugosity index calculations (see Table 1\autoref{tab:1}). Additionally, **pyTopoComplexity** 
includes a module for calculating the Terrain Position Index (TPI), a widely used metric that is considered effective 
in geomorphology research [@Newman2018; @Deumlich2010; @Liu2011]. The TPI is also often used alongside rugosity index 
calculations in marine geological and ecological analyses [@Wilson2007; @Walbridge2018].

This toolkit can detect the grid spacing and unit of the projected coordinate system (acceptable in meters, 
U.S. survey feet, and international feet) from the input raster DTM file (GeoTIFF format) and 
automatically conduct unit conversions in necessary calculation steps to ensure data consistency and 
reproducibility. Results at nodes affected by edge effects due to no-data values outside the input raster 
will be removed by default. Users can define the suitable spatial scale to match their research purposes 
and choose computational approaches (e.g., chunk processing, faster mathematical approximations) to 
optimize performance (see details in the **Methods and features overview** section).

| Modules  | Classes | Method Descriptions | References |
| ------------- | ------------- | ------------- | ------------- |
| pycwtmexhat.py | CWTMexHat | Quantifies the wavelet-based curvature of the land surface using two-dimensional continuous wavelet transform (2D-CWT) with a Mexican Hat wavelet | @Booth2009; @Booth2017 |
| pyfracd.py | FracD | Conducts fractal dimension analysis on the land surface using variogram procedures | @Wen1997; @PardoIguzquiza2020|
| pyrugostiy.py | RugosityIndex | Calculates the Rugosity Index of the land surface | @Jenness2004; @DuPreez2015 |
| pytpi.py | TPI | Calculate Terrain Position Index of the land or seafloor surface | @Weiss2001; @Walbridge2018 |
Table: Table 1 \label{tab:1}: Modules contained in the **pyTopoComplexity** package.

Each module of the **pyTopoComplexity** is provided with a corresponding [example Jupyter Notebook file](https://github.com/GeoLarryLai/pyTopoComplexity/tree/main/example) 
for usage instructions, using the Light Detection and Ranging (LiDAR) DTM data of a deep-seated landslide that 
occurred in 2014 in the Oso area of the North Fork Stillaguamish River valley, Washington State, USA 
[@WashingtonGeologicalSurvey2023]. In the software repository, we also include an additional Jupyter Notebook file 
[**nonlineardiff_Landlab.ipynb**](https://github.com/GeoLarryLai/pyTopoComplexity/blob/main/example/nonlineardiff_Landlab.ipynb), 
which allows researchers to simulate the smoothing of topography over time via terrestrial nonlinear hillslope 
diffusion processes [@Roering1999]. This is achieved by employing the `TaylorNonLinearDiffuser` module from the 
`terrainBento` Python package [@Barnhart2019] and running the simulation in the **Landlab** environment [@Hobley2017].

@@

The Mexican hat wavelet is proportional to the second derivative of a Gaussian envelope, with its Fourier wavelength ($\lambda$) 
dependent on the grid spacing ($\Delta$) of the input DTM raster. The wavelet function $\psi$ is scaled according to the 
wavelet scale parameter $s$ and the grid spacing $\Delta$, ensuring that the resultant wavelet coefficient $C$ signifies 
concave and convex landforms corresponding to the wavelet scale $s$. Users can define the value of $\lambda$ in meters as the 
targeted spatial scale for landform roughness analysis, and the **pycwtmexhat.py** module will automatically compute the wavelet 
scale $s$ based on the grid spacing ($\Delta$) of the input raster file (Figure 1)\autoref{fig:1}.

@@

, where $TD$ is the topological dimension in the Euclidean space of the fractional Brownian motion. For one-dimensional fractional Brownian 
motion, $TD = 1$. The fractal dimension of the two-dimensional surface $(FD)_2^*$ can be estimated as the average fractal dimension of 
the one-dimensional profiles $(FD)_1^*$:

@@

## Rugosity Index calculation

The **pyrugosity.py** module in **pyTopoComplexity**  measures the Rugosity Index of the land surface, which is widely used to assess structural 
complexity of the topography. Such method has been applied in classifying seafloor types by marine geologists and geomorphologist, understanding small-scale 
hydrodynamics by oceanographers, and studying available habitats in the landscape by ecologists and coral biologists [@Lundblad2006; @Wilson2007].

The Rugosity Index is determined as the ratio of the contoured area (i.e., true geometric surface area) to the planimetric area within the square 
moving window, highlighting smaller-scale variations in surface height:

@@

This module adapts the Triangulated Irregular Networks method from @Jenness2004 to approximate the contoured area as the sum of 
eight truncated-triangle areas. These triangles connect the central grids, four corner grids, and four grids at the middle points of the 
surrounding edges within the moving window. If no local slope correction is applied, the planimetric area is considered to be the horizontal 
planar area of the moving window, as described in @Jenness2004. Another approach considers slope correction where to the planimetric area is 
projected onto an plane of the local gradient [@DuPreez2015].

By definition, the Rugosity Index is as a minimum value of one (completely flate surface). Typical valuesrange of the conventional rugosity 
index (without slope correction) from one to three although larger values are possible in very steep terrains. The slope-corrected rugosity 
index, also called arc-chord ratio (ACR) Rugosity Index, could provide a better representation of local surface complexity (Figure 3)\autoref{fig:3}.

![Figure 3. Calculated results of the Arc-Chord Ratio Rugosity Index (upper) and the conventional Rugosity Index (lower) for the 2014 Oso Landslide 
region, using the **pyrugosity.py** module. The calculations were performed with different moving window sizes: 17 by 17 grids (left), 49 by 49 grids 
(middle), and 83 by 83 grids (right). Grid spacing is 3 U.S. survey feet (approximately 0.9144 meters). \label{fig:3}](fig3-pyrugosity.png){ width=100% }

## Terrain Position Index calculation

The **pytpi.py** module in **pyTopoComplexity** calculates the Terrain Position Index (TPI) of the land surface. The TPI, also known as the 
Topographic Position Index in terrestrial studies [@Weiss2001], measures the relative topographic position of a point compared to its 
surrounding landforms. This metric highlights regions that are relatively higher or lower than their surroundings, which is useful for 
distinguishing landscape features such as hilltops, valleys, flat plains, and slopes. In oceanography, an equivalent metric is the 
Bathymetric Position Index (BPI), which applies the TPI algorithm to bathymetric data to evaluate seafloor complexity.

TPI is widely applicable for various purposes, including determining surface ruggedness [@Newman2018], classifying terrain [@zwolinski2015], 
assessing local soil formation and hydrodynamics [@Deumlich2010; @Liu2011], and identifying habitat hotspots [@Wilson2007]. It is calculated 
by comparing the elevation of a grid cell ($Z_{\text{grid}}$) to the mean elevation of its surrounding grid cells ($Z_{\text{neighborhood}}$) 
within a specified neighborhood:

$$
TPI = Z_{\text{grid}} - \text{mean}(Z_{\text{neighborhood}})
$$

In this module, the TPI value is computed at the central grid within a square-shaped moving window. Users can specify the size of the moving 
window (i.e., the number of grids along each edge) to assess the topographic position at different spatial scales. Positive TPI values 
indicate elevated features, negative values represent depressions, and values close to zero denote flat or gently sloping areas. The **pytpi.py** 
module also returns the absolute values of the TPI (Figure 4\autoref{fig:4}). This absolute TPI raster is suggested to be effective for 
quick assessments in mapping and relative dating of deep-seated landslide deposits using high-resolution DTM data [@Lai2023].

![Figure 4. Calculated results of the Terrain Position Index (upper) and its absolute values (lower) for the 2014 Oso Landslide region, 
using the **pytpi.py** module. The calculations were performed with different moving window sizes: 17 by 17 grids (left), 49 by 49 grids 
(middle), and 83 by 83 grids (right). Grid spacing is 3 U.S. survey feet (approximately 0.9144 meters). \label{fig:4}](fig4-pytpi.png){ width=100% }

@@

The [**nonlineardiff_Landlab.ipynb**](https://github.com/GeoLarryLai/pyTopoComplexity/blob/main/example/nonlineardiff_Landlab.ipynb) notebook 
in the **pyTopoComplexity** package offers a sophisticated tool for simulating landscape evolution 
through nonlinear diffusion processes due to near-surface soil disturbances and downslope sediment creep [@Roering1999]. This tool runs the 
simulation in the **Landlab** environment (version >= 2.7) [@Hobley2017] with the `TaylorNonLinearDiffuser` module from the `terrainBento` 
Python package [@Barnhart2019]. The main simulation iteratively applies the nonlinear diffusion model to predict changes in surface 
elevation $z$ over time $t$:

@@

This notebook provides a comprehensive workflow that guides users through setting up, importing raster files, and running simulations. Since 
**Landlab** primarily handles DTM data in ESRI ASCII format, this notebook includes utility functions for converting raster files between GeoTIFF 
and ESRI ASCII formats. Users are required to specify the values for $S_c$, $K$, the length of each time step in years, and the final time to 
stop the simulation. The example included in the notebook uses LiDAR DTM data from the 2014 Oso Landslide [@WashingtonGeologicalSurvey2023], 
with parameter values provided in @Booth2017, in an attempt to reproduce the simulation results presented in that study (Figure 5\autoref{fig:5}).

![Figure 5. Hillshade maps of the 2014 Oso Lanslide region and surface smoothing evolution over 15,000 years predicted by a 
nonlinear hillslope diffusion model used in the [**nonlineardiff_Landlab.ipynb**](https://github.com/GeoLarryLai/pyTopoComplexity/blob/main/example/nonlineardiff_Landlab.ipynb) 
notebook, in attempt to reproduce the simulation results in @Booth2017.\label{fig:5}](fig5-smoothing.png){ width=100% }

@@

The development of **pyTopoComplexity** is part of a collaborative effort within the [Landslide subteam](https://cascadiacopeshub.org/team-1-landslides/) of the 
[Cascadia Coastlines and Peoples Hazards Research Hub (Cascadia CoPes Hub)](https://cascadiacopeshub.org), funded by the National Science Foundation Award ([#2103713](https://www.nsf.gov/awardsearch/showAward?AWD_ID=2103713&HistoricalAwards=false)). 
We extend special thanks to Dr. Eulogio Pardo-Igúzquiza, who generously shared his Fortran code for fractal dimension analysis used in his work 
[@PardoIguzquiza2022b], which  inspired the development of the **pyfracd.py** module.
